---
title: "Molecular profiling of glioblastoma: spatial comparison against control brain tissue"
author: "Chiara Falleroni & Luca Marzilli"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: False
    number_sections: true
    fig_width: 8
    fig_height: 6
    highlight: tango
    df_print: kable
  html_document: 
    toc: true
    toc_float: true
    theme: cosmo
    number_sections: true
    fig_width: 8
    fig_height: 6
editor_options: 
  markdown: 
    # chiedi a chiara se le piaciono gli output inline.
    wrap: 80
    # chunk_output_type: console 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
```


\newpage
\tableofcontents
\newpage

```{r}
library(RColorBrewer)
library(omicsdata)
library(SummarizedExperiment)
library(tidyverse)
library(tibble)
library(edgeR)
library(limma)
library(DESeq2)
library(EDASeq)
#install.packages("factoextra")
library(factoextra)
#BiocManager::install("ComplexHeatmap")
library(ComplexHeatmap)
library(biomaRt)
library(pheatmap)
```

```{r}
Class_pal2way = c("control"="turquoise4", "glioma"="firebrick2")
Class_pal3way = c("control"="turquoise4", "glioma_CE"="darkred", "glioma_NE"="salmon2")
continuous_pal2 = colorRampPalette(c("white","plum1","purple4"))(100)
continuous_pal3 = colorRampPalette(c("springgreen4", "antiquewhite1", "orangered2"))(100)
```

# Biological context and study objectives 

This work is based on the study by Gill et al., 2014, which investigates the molecular and transcriptional landscape of glioblastoma using RNA sequencing approaches. The article provides a comprehensive framework for understanding the complexity of this tumor by integrating high-throughput sequencing data with spatial and biological context, offering valuable insights into the heterogeneity and cellular composition of glioblastoma.  

In this article, ***glioblastoma (GBM)*** is described as the most aggressive primary brain tumor and is characterized by extensive intratumoral heterogeneity and diffuse infiltration into surrounding brain tissue.  

Due to this infiltrative behavior, complete surgical resection is not possible, and residual tumor cells located in the margins of the tumor inevitably contribute to disease recurrence. 
Recent transcriptomic studies have shown that glioblastoma can be classified into distinct molecular subtypes, including proneural, classical, and mesenchymal.  

However, most of these classifications are based on samples derived from the **contrast-enhancing (CE)** tumor core, which corresponds to regions that appear enhanced in MRI scans due to the presence of a disrupted blood-brain barrier and are typically enriched in highly proliferative tumor cells. This focus on the tumor core may overlook critical aspects of tumor biology, particularly those related to invasion and interaction with the surrounding microenvironment.  

In contrast, **non-enhancing (NE)** tumor regions refer to the infiltrative margins of the tumor that do not show contrast enhancement in imaging and contain a heterogeneous mixture of neoplastic cells and normal brain cells, such as neurons and glial cells. These regions are thought to play a crucial role in tumor progression and recurrence, as they harbor residual tumor cells that evade surgical resection and may exhibit distinct transcriptional programs compared to the tumor core.  

In this study, we analyze RNA-seq data from glioma and non-neoplastic brain tissues to investigate differences in gene expression profiles. Specifically, we aim to compare tumor samples with normal brain tissue and to further explore differences between CE and NE tumor regions. This allows us to assess how cellular composition and tumor heterogeneity influence gene expression patterns.

\vspace{2cm}

The main objectives of this analysis are:  

- to identify genes differentially expressed between glioma and non-neoplastic tissues;  
- to evaluate the extent to which tumor samples differ from normal brain tissue at the transcriptomic level;  
- to investigate how different tumor regions (CE vs NE) contribute to variability in gene expression;  
- to interpret the results in light of cellular heterogeneity and tumor microenvironment.



# Experimental design 

## Data provenance: original article's methods

### Physical collection of samples and replicates

The researchers analyzed 69 adult patients who were presenting for open surgical resection of a glioma. They also collected 17 samples of healthy (non neoplastic) brain tissue from 11 patients undergoing surgery for other reasons (such as hydrocephalus or epilepsy) to use as a baseline for comparison. Having so little patients against so many samples obviously raises statistical independence problems.

Before beginning to remove the main bulk of the tumor (surgical debulking), the neurosurgeons took small fragments of tissue (measuring approximately 1.0 x 0.5 x 0.5 cm). To know exactly where they were taking the samples from, they used an intraoperative neuro-navigation system (Brainlab) based on MRI scans. They specifically targeted two areas: the "core" of the tumor (the contrast-enhancing area, called "CE") and the infiltrative margins of the tumor (the non-enhancing area, visible on FLAIR sequences, called "NE").

Once taken, directly in the operating room, each piece of tissue was divided in half: one piece was immediately flash-frozen in liquid nitrogen (to preserve the RNA unaltered), while the other piece was immersed in 10% formalin to "fix" and preserve the tissue structure for microscopic analysis.

### Histological analysis 

The formalin-fixed samples were cut into very thin slices, stained, and analyzed under a microscope to visually evaluate cellular density, the presence of abnormalities (like necrosis or blood vessel proliferation), and to physically mark and count the remaining entrapped healthy neurons (using a marker called NeuN).

### RNA-sequencing

Genome-wide expression profiles were generated by performing RNA-sequencing on a total of 92 flash-frozen samples. These included 39 from the tumor core (CE), 36 from the infiltrative margins (NE), and 17 from healthy brain tissue (NB).

For each sample, sequencing "libraries" were prepared using the Illumina TruSeq v2 RNA-Seq kit. The actual sequencing was performed on an Illumina HiSeq 2000 machine, which produced between 15 and 30 million "single-end, 100-base reads" per sample.

Alignment was performed via the software Tophat 2, and expression levels were calculated with Cufflinks 2, using the FPKM metric (Fragments Per Kilobase per Million mapped reads).

### Original author's bioinformatic pipeline

Using the RNA-seq data, they calculated a mathematical correlation (Spearman) between the genetic profiles of their samples and those of a massive public reference database (TCGA). This was used to classify each region of the tumor into one of the known molecular subtypes (Proneural, Classical, Mesenchymal, or Neural).

Subsequently, the authors conducted a **Computational deconvolution**: a mathematical practice performed in order to "untangle" the signal from tumor margins; in fact this samples where often a complex mixture (see the previous paragraph) of tumor and non-tumor brain cells.  
Before the deconvolution the most of the NE tumor samples were assigned to the "neural" TCGA expression profile; but this stopped being the case after mathematical unmixture.   
The authors therefore implicitly demostrated how the neural subtype was not an actual expression profile for the formerly analyzed gliomas, but an artifact caused by cellular contamination.  

## Dataset structure

The gene expression data related to the study were imported into the analysis environment as a SummarizedExperiment object (originating from the SRP044668 project), a standard data structure in Bioconductor designed to securely coordinate expression matrices with their respective clinical and technical metadata. 

The dataset consists of a single expression matrix, named raw_counts, which contains the unnormalized raw read counts. This matrix is substantial in size: it contains expression values for 63,856 genes (annotated for Homo sapiens) measured across a set of 94 biological samples.  
Alongside the data matrix, the object integrates an extensive annotation and metadata table (colData) consisting of 175 variables for each sample.

# Data and preprocessing 

## Data loading and a brief look at it

```{r}
if (file.exists("SRP044668.rds")) {
  SRP = readRDS("SRP044668.rds")
} else {
  stop("dear user: the necessary data is not present in the current working directory or 
        does not have the right name, please check.")
}
```

Let's give a look at the data's structure. 

```{r}
#str(SRP)

cat("The dataset contains", nrow(SRP), "genes across", ncol(SRP), "samples.")
```

```{r}
cat("The object contains the assay:", names(assays(SRP)))
```

## Data subsetting

We are now going to subset the data coherently with our goals.

```{r}
#sample_info = colData(SRP)$sra.sample_attributes 

# re-ordering that annoying last observation with the other glioma_NE
nuovo_ordine <- c(94, 1:93)
SRP = SRP[, nuovo_ordine]

sample_info = colData(SRP)$sra.sample_attributes

# Using new columns in the ColData I classify the columns (samples) of the 
# dataset based on the type.

# 3-way classification

colData(SRP)$class_3way =
  case_when(grepl("brain tissue", sample_info)~"control",
            grepl("contrast-enhancing", sample_info)~"glioma_CE",
            grepl("non-enhancing", sample_info)~"glioma_NE")

# 2-way classification

colData(SRP)$class_2way =
  case_when(grepl("brain tissue", sample_info)~"control",
            grepl("glioma", sample_info)~"glioma")

# FACTORIALIZATION (necessary for graphs and other procedures)

colData(SRP)$class_3way = factor(colData(SRP)$class_3way, 
                                  levels = c("control", "glioma_CE", "glioma_NE"))

colData(SRP)$class_2way = factor(colData(SRP)$class_2way, 
                                  levels = c("control", "glioma"))
```

## Unexpressed genes removal

As a team, we decide to exclude genes with expression levels at 0 or extremely low levels in all samples here, before the EDA, and not in the quality control paragraph. We do this because we believe this row filtering is not a quality control but a basic definition of the rows we consider "data". We also decide to use the FilterByExpr (from edgeR) function, a much more statistically reliable tool.

```{r}
# genes with all zero counts
cat(
  "Genes with all zero counts:",
  sum(rowSums(assay(SRP, "raw_counts")) == 0),
  "\n"
)

# lowly expressed genes
cat(
  "Genes with total counts lower than 15:",
  sum(rowSums(assay(SRP, "raw_counts")) < 15),
  "\n"
)

# missing values
cat(
  "Missing values in the count matrix:",
  sum(is.na(assay(SRP, "raw_counts"))),
  "\n"
)

# duplicated genes
cat(
  "Duplicated gene identifiers:",
  sum(duplicated(rownames(SRP))),
  "\n"
)

# dimensions before filtering
cat(
  "Dataset dimensions before filtering:",
  nrow(SRP),
  "genes and",
  ncol(SRP),
  "samples.\n"
)

# filtering
keep = filterByExpr(
  assay(SRP, "raw_counts"),
  group = SRP$class_3way
)

SRP = SRP[keep, ]

# dimensions after filtering
cat(
  "Dataset dimensions after filtering:",
  nrow(SRP),
  "genes and",
  ncol(SRP),
  "samples."
)
```

# Pre-normalization EDA 

Now, before proceeding, we will use some graphical tools to investigate the general nature of our counts. This analysis is going to be helpful in guiding us through the right choices in the next section: data normalization and quality control.

## Analysis of the genetic expression distribution in the samples

```{r}

meta_df = data.frame(Sample=colnames(SRP), Class2way=SRP$class_2way, Class3way=SRP$class_3way)

dfCounts_long = as.data.frame(assay(SRP, "raw_counts")) |>
  rownames_to_column(var="Gene") |>
  pivot_longer(cols=-Gene, names_to = "Sample", values_to = "Expression")

dfCounts_long = dfCounts_long |> left_join(meta_df, by = "Sample")

#dfCounts_long |>
#  ggplot( aes(x=Sample, y=Expression, fill = Class2way)) +
#  geom_boxplot() +
#  scale_fill_manual(values=Class_pal2way) +
#  theme(
#    axis.text.x = element_blank(),  
#    axis.ticks.x = element_blank(), 
#    axis.title.x = element_blank()  
#  )

boxplot(assay(SRP, "raw_counts"),
        las = 2,
        col=Class_pal2way[colData(SRP)$class_2way],
        #ylab = "Expression (raw)",
        main = "Boxplot of raw counts")

  
```
This are boxplots of the expression levels of some randomly chosen samples. As seen in the laboratories, we clearly need a logarithmic transformation.

```{r}
dfCounts_long$Expression = log1p(dfCounts_long$Expression)

#dfCounts_long |>
#  ggplot( aes(x=Sample, y=Expression, fill = Class2way)) +
#  geom_boxplot() +
#  scale_fill_manual(values=Class_pal2way) +
#  theme(
#    axis.text.x = element_blank(),  
#    axis.ticks.x = element_blank(), 
#    axis.title.x = element_blank())

# creation of a new log1p assay in the SRP summarized experiment object
assays(SRP)$logcounts <- log1p(assay(SRP, "raw_counts"))

boxplot(assay(SRP, "logcounts"),
        #outline = FALSE,
        las = 2,
        col=Class_pal2way[colData(SRP)$class_2way],
        ylab = "Expression (log1p)",
        main = "Boxplot after log1p transformation")

```
Now the distributions are much better.

Here we show the distribution densities conditioned on the sample class of the expressions.

```{r}
dfCounts_long |>
  ggplot(aes(x=Expression, fill=Class2way, colour = Class2way)) +
        geom_density(alpha=0.4) +
        scale_fill_manual(values = Class_pal2way) +
        scale_color_manual(values = Class_pal2way)

dfCounts_long |>
  ggplot(aes(x=Expression, fill=Class2way, colour = Class3way)) +
        geom_density(alpha=0.4) +
        scale_fill_manual(values = Class_pal3way) +
        scale_color_manual(values = Class_pal3way)
```

## MD, MA and RLE plots

### MA

```{r}
limma::plotMA(assay(SRP, "logcounts"))
```

Here the limma's plotMA() function is working as a quality control tool. It's confronting the first sample with the mean of the others, and the asymmetries clearly comunicate us the need for normalization.

### MD

```{r}
EDASeq::MDPlot(assay(SRP, "raw_counts"), c(6, 7))
```

### RLE

```{r}
plotRLE(assay(SRP,"raw_counts"), outline = FALSE, las = 2,
        ylab = "RLE", main = "RLE of raw counts", col = Class_pal3way[colData(SRP)$class_3way])
```

Both the MD and RLE plots strengthen what was already visible in the MA: the data need normalization.

## PCA

```{r}
pca = prcomp(t(assay(SRP, "logcounts")))
screeplot(pca)

fviz_pca_ind(pca,
             geom.ind = "point",       
             pointsize = 1,            
             col.ind = SRP$class_3way,      
             palette = Class_pal3way,       
             addEllipses = TRUE,       
             ellipse.type = "convex")
```

Lastly, we here performed and plotted a pca on the un-normalized data. 

## GC content bias evaluation  
GC content can introduce biases in library preparation and sequencing, potentially affecting the quantification of gene expression.
It's important to check GC content bias for eventually filtering out sequences.  
First, we need to retrieve the amount of *GC* for each gene. To do this we use the **rowRanges** row variables and the **getGeneLengthAndGCContent** function of the *EDASeq* package.

```{r}
#Gene ID extraction and annotation retrieval
gene_ids = as.character(rowRanges(SRP)$gene_id)
gene_ids = sub("\\..*$", "", gene_ids)

gc_info = getGeneLengthAndGCContent(
  id = gene_ids,
  org = "hg38",
  mode = "org.db"
)
```

Gene identifiers were extracted from the dataset and processed to remove version numbers. GC content and gene length information were then retrieved using EDASeq annotation functions based on the hg38 reference genome.

```{r}
#Data cleaning and alignment
gc_info = gc_info[!is.na(gc_info[, 2]), ]

valid_genes = gene_ids %in% rownames(gc_info)

SRP_filtered = SRP[valid_genes, ]
gene_ids = gene_ids[valid_genes]
```

Genes lacking GC content information were removed.  
The dataset was then aligned with the annotation data to retain only genes with available GC information, ensuring consistency between expression data and gene features.

```{r}
#Count matrix preparation
counts_matrix = assay(SRP_filtered, "raw_counts")

rownames(counts_matrix) = gene_ids
```

The raw count matrix was extracted and gene identifiers were standardized to ensure consistency with the annotation dataset.

```{r}
#Removal of duplicated genes
dup_genes = duplicated(rownames(counts_matrix))

counts_matrix = counts_matrix[!dup_genes, ]
gc_info = gc_info[!dup_genes, ]
SRP_filtered = SRP_filtered[!dup_genes, ]
```

Duplicated gene identifiers, arising from multiple transcript mappings, were removed to ensure a one-to-one correspondence between genes and expression values.

```{r}
#Construction of SeqExpressionSet
seq_set = newSeqExpressionSet(
  counts = counts_matrix,
  phenoData = AnnotatedDataFrame(
    data.frame(
      condition = SRP_filtered$class_3way,
      row.names = colnames(counts_matrix)
    )
  ),
  featureData = AnnotatedDataFrame(
    data.frame(
      gc_content = gc_info[, 2],
      gene_length = gc_info[, 1]
    )
  )
)
```

A SeqExpressionSet object was constructed to integrate expression data with sample metadata and gene-level features, including GC content and gene length, enabling downstream bias analysis.

```{r}
#GC bias plot
biasPlot(seq_set, "gc_content", log = TRUE)
```
The GC bias plot shows a clear non-linear relationship between gene expression and GC content across all samples. Expression levels increase at intermediate GC values and decrease at both low and high GC content, indicating the presence of GC-dependent technical bias. The consistency of this pattern across samples suggests that the bias is systematic rather than sample-specific.

GC content bias was evaluated as part of the quality assessment to investigate potential gene-specific technical effects. Although a GC-dependent trend was observed, normalization was subsequently performed using DESeq2, which accounts for major sources of between-sample variability. Therefore, GC bias was not explicitly corrected but considered in the interpretation of the results.


# Data normalization and quality control

Although GC-dependent bias was observed during exploratory analysis, normalization was performed using DESeq2, which accounts for major sources of between-sample variability. Additional within-sample normalization was not applied to maintain consistency with the statistical modeling framework.


## Data extraction and DESeq2 dataset construction

```{r}
counts_matrix = assay(SRP_filtered, "raw_counts")
metadata = as.data.frame(colData(SRP))

dds = DESeqDataSetFromMatrix(
  countData = counts_matrix,
  colData = metadata,
  design = ~ class_3way
)

```

A DESeq2 dataset object was created by combining the raw count matrix with sample metadata and specifying the experimental design based on the class_3way variable.  

## DESeq2 normalization

```{r}
dds = DESeq(dds)
```

Normalization was carried out within the DESeq2 framework through the estimation of size factors, correcting for differences in sequencing depth and compositional biases across samples.

## Variance stabilizing transformation (VST)
```{r}
vsd = vst(dds, blind = FALSE)
```

A variance stabilizing transformation (VST) was then applied to reduce the dependence of variance on mean expression levels, making the data more suitable for exploratory analyses and visualization such as PCA and clustering. The VST is much more robust to distortions than simple log1p transformation.


# Post-normalization EDA

Post-normalization quality control is performed to assess the effectiveness of the normalization procedure. In particular, boxplots and Relative Log Expression (RLE) plots are used to evaluate the distribution of expression values across samples and to verify the reduction of technical variability. 
These visualizations allow us to determine whether samples are comparable and suitable for downstream analyses.

## Boxplot

```{r}
boxplot(assay(vsd),
        #outline = FALSE,
        las = 2,
        col=Class_pal2way[colData(vsd)$class_2way],
        ylab = "Expression (VST)",
        main = "Boxplot after normalization")


meta_df = data.frame(Sample=colnames(vsd), Class2way=SRP$class_2way, Class3way=SRP$class_3way)

dfCounts_long = as.data.frame(assay(vsd)) |>
  rownames_to_column(var="Gene") |>
  pivot_longer(cols=-Gene, names_to = "Sample", values_to = "Expression")

dfCounts_long = dfCounts_long |> left_join(meta_df, by = "Sample")

#dfCounts_long |>
#  ggplot( aes(x=Sample, y=Expression, fill = Class2way)) +
#  geom_boxplot() +
#  scale_fill_manual(values=Class_pal2way) +
#  theme(
#    axis.text.x = element_blank(),  
#    axis.ticks.x = element_blank(), 
#    axis.title.x = element_blank()  
# )

```
Boxplots of variance-stabilized data show highly consistent expression distributions across samples after normalization.  
The alignment of medians and the comparable spread of values indicate that technical variability has been effectively reduced, making the samples suitable for downstream comparative analysis.

## Density plot

```{r}
dfCounts_long |>
  ggplot(aes(x=Expression, fill=Class2way, colour = Class2way)) +
        geom_density(alpha=0.4) +
        scale_fill_manual(values = Class_pal2way) +
        scale_color_manual(values = Class_pal2way) +
        labs(title = "Density Plot after VST Normalization")
```
The density plot of VST-normalized data shows a characteristic shift of unexpressed genes to a baseline value of approximately 9. This offset is the expected mathematical behavior of the Variance Stabilizing Transformation, which stabilizes the variance at the lower end of expression. Most importantly, the perfect overlap of the density curves across the experimental groups confirms that technical biases have been successfully mitigated, ensuring the samples are robustly comparable for downstream analyses.

## RLE plot 

```{r}
plotRLE(assay(vsd),
        outline = FALSE,
        las = 2,
        col=Class_pal3way[colData(vsd)$class_3way],
        ylab = "RLE",
        main = "RLE after normalization")
```
The RLE plot shows values centered around zero with a consistent spread across samples, indicating that normalization effectively reduced technical variability.
A small number of samples display slight deviations from the overall trend, but these remain limited and do not suggest the presence of strong outliers or residual technical bias. Overall, the data appear well-balanced and suitable for downstream analysis.

Now we perform Principal Component Analysis (PCA) on variance-stabilized data to investigate the main sources of variation within the dataset and to evaluate the separation between biological groups. This analysis enables visualization of sample clustering patterns and helps assess whether normalization preserves biologically meaningful differences.

## PCA

```{r}
pca = prcomp(t(assay(vsd)))
screeplot(pca)

fviz_pca_ind(pca,
             geom.ind = "point",       
             pointsize = 1,            
             col.ind = vsd$class_3way,      
             palette = Class_pal3way,       
             addEllipses = TRUE,       
             ellipse.type = "convex")
```
The PCA plot shows a partial separation between glioma and control samples along the first principal component, indicating that biological differences contribute significantly to the overall variability.  
Control samples tend to cluster more tightly, while glioma samples display greater dispersion, reflecting higher heterogeneity within the tumor group. A certain degree of overlap is observed, likely due to the presence of non-neoplastic components within glioma samples or infiltrative tumor regions.  Overall, the distribution of samples suggests that normalization was effective and that biologically relevant variation is captured by the principal components.

Due to the Screeplot similar PC2 and PC3 variances we also plot here in the post-normalization EDA a graph made with PC1 and PC3 as axes.

```{r}
fviz_pca_ind(pca,
             geom.ind = "point", 
             axes = c(1,3),
             pointsize = 1,            
             col.ind = vsd$class_3way,      
             palette = Class_pal3way,       
             addEllipses = TRUE,       
             ellipse.type = "convex")
```
Interestingly, the PC3 axis spreads the data points quite similarly to the PC2 (they probably go alongside, even if they are independent by definition). Let's try to plot the PC2 and the PC3 together then.

```{r}
fviz_pca_ind(pca,
             geom.ind = "point", 
             axes = c(2,3),
             pointsize = 1,            
             col.ind = vsd$class_3way,      
             palette = Class_pal3way,       
             addEllipses = TRUE,       
             ellipse.type = "convex")
```
Actually, the PC2 goes approximately alongside the PC3 only for control and glioma_NE points, In glioma_CE they seem to be inversely correlated (even if, as already said, they must be orthogonal by definition when we look at the whole data points)!

### Consulting the PCA runes: a forecast of DEGs

In this paragraph we aim to get a first glance at our dataset transcriptomical variability using the PCs loadings

```{r}
pc1_loadings = pca$rotation[,1]
sorted_pc1_loadings = pc1_loadings[order(abs(pc1_loadings), decreasing = TRUE)]
pc1_var_drivers_df = data.frame(names(sorted_pc1_loadings[1:10]), as.numeric(sorted_pc1_loadings[1:10]))
colnames(pc1_var_drivers_df) = c("Gene", "Contribution to PC1")

pc1_var_drivers_df$Gene_clean <- gsub("\\..*", "", pc1_var_drivers_df$Gene)

mart <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")

gene_symbols <- getBM(
  attributes = c("ensembl_gene_id", "external_gene_name"),
  filters = "ensembl_gene_id",
  values = pc1_var_drivers_df$Gene_clean,
  mart = mart
)

pc1_var_drivers_df =
  pc1_var_drivers_df |> 
  left_join(gene_symbols, by = c("Gene_clean" = "ensembl_gene_id"))

colnames(pc1_var_drivers_df) = c("Ensembl ID", "Contribution to PC1", "Clean ensembl ID", "Gene name")
pc1_var_drivers_df = pc1_var_drivers_df[, c(4, 1, 3, 2)]

pc1_var_drivers_df

```

This 10 genes and their contribution to PC1 are sending us a clear hint: the most veriability in our dataset is encoded in the loss of neurobiological functions. The genes are all neurobiology related (**Receptors**: *GRIN1*, *GABRA1* ; **vescicular traffic**: *SLC17A7*, *SYT1*,... ; **Membrane and cytoskeleton**: *PACSIN1*, *NEFL* ; **Intracellular signaling**: *CAMK2A*, *VSNL1* ), and the less expressed they are, the more on the right of the graph (positive values) the samples are going to be, alongside the glioma_CE samples. **What we can state now then, is that tumor glioma cells tend with all probabilities to loose the expression of neuronal functions.**


```{r}
pc2_loadings = pca$rotation[,2]
sorted_pc2_loadings = pc2_loadings[order(abs(pc2_loadings), decreasing = TRUE)]
pc2_var_drivers_df = data.frame(names(sorted_pc2_loadings[1:10]), as.numeric(sorted_pc2_loadings[1:10]))
colnames(pc2_var_drivers_df) = c("Gene", "Contribution to PC2")

pc2_var_drivers_df$Gene_clean <- gsub("\\..*", "", pc2_var_drivers_df$Gene)

gene_symbols <- getBM(
  attributes = c("ensembl_gene_id", "external_gene_name"),
  filters = "ensembl_gene_id",
  values = pc2_var_drivers_df$Gene_clean,
  mart = mart
)

pc2_var_drivers_df =
  pc2_var_drivers_df |> 
  left_join(gene_symbols, by = c("Gene_clean" = "ensembl_gene_id"))

colnames(pc2_var_drivers_df) = c("Ensembl ID", "Contribution to PC2", "Clean ensembl ID", "Gene name")
pc2_var_drivers_df = pc2_var_drivers_df[, c(4, 1, 3, 2)]

pc2_var_drivers_df

```

This new bundle of genes is now completely composed of oligodentrocytes' (and partially astrocites') associated genes. We believe it's separating samples with regard to their cellular composition (neurons vs glia cells proportion).

```{r}
pc3_loadings = pca$rotation[,3]
sorted_pc3_loadings = pc3_loadings[order(abs(pc3_loadings), decreasing = TRUE)]
pc3_var_drivers_df = data.frame(names(sorted_pc3_loadings[1:10]), as.numeric(sorted_pc3_loadings[1:10]))
colnames(pc3_var_drivers_df) = c("Gene", "Contribution to PC3")

pc3_var_drivers_df$Gene_clean <- gsub("\\..*", "", pc3_var_drivers_df$Gene)

gene_symbols <- getBM(
  attributes = c("ensembl_gene_id", "external_gene_name"),
  filters = "ensembl_gene_id",
  values = pc3_var_drivers_df$Gene_clean,
  mart = mart
)

pc3_var_drivers_df =
  pc3_var_drivers_df |> 
  left_join(gene_symbols, by = c("Gene_clean" = "ensembl_gene_id"))

colnames(pc3_var_drivers_df) = c("Ensembl ID", "Contribution to PC3", "Clean ensembl ID", "Gene name")
pc3_var_drivers_df = pc3_var_drivers_df[, c(4, 1, 3, 2)]

pc3_var_drivers_df

```

Lastly, we encounter a clearly sex-driven PC. Positive genes (*XIST*, *TSIX*,...) are female transcripts, while negatives (*DMRT2*,...) are males. In spite of this there are other genes with non sex-related functions such as Lactotransferrin or SOX10. We suppose this might be linked to dataset imbalance between sex and tumor rapresentation (e.g. all SOX10 upregulated tumors were coming from females) but that is not necessarily the case; the fact the sample population came from only 27 different patients makes this biases really easy to appear (please find the article's supplementary methods in the zip file sent, table S1).

### Sex and the plot

Here we aim to confirm the sex separation PC3 might be doing what we supposed in the previous paragraph. Before starting, we had to set a M/F ground truth but this has not been easy since we didn't find a downloadable CSV (or any tabular file) of the patient's table (table S1), and this PDF table was moreover lacking any form of information to be aligned to our 94 brain samples. We therefore exploited the percentual amount of Y chromosome reads. 

```{r}
df_qc <- data.frame(
  chrY_reads = colData(SRP)$`recount_qc.aligned_reads%.chry`
)

ggplot(df_qc, aes(x = chrY_reads)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "black") +
  geom_vline(xintercept = 0.035, color = "red", linetype = "dashed", size = 1) +
  theme_minimal() +
  labs(
    title = "chrY reads distribution",
    x = "Reads on chrY (%)",
    y = "number of samples"
  )

estimated_sex = ifelse(
  colData(SRP)$`recount_qc.aligned_reads%.chry` > 0.035,
  "male",
  "female"
)

```

```{r}
summary(colData(SRP)$`recount_qc.aligned_reads%.chry`)
```


```{r}
table(estimated_sex, colData(SRP)$class_2way)
```

```{r}
table(estimated_sex, colData(SRP)$class_3way)
```



```{r}
estimated_sex = factor(
  estimated_sex,
  levels = c("female", "male")
)

fviz_pca_ind(
  pca,
  geom.ind = "point",
  axes = c(1,3),
  pointsize = 2,
  col.ind = estimated_sex,
  palette = c("hotpink", "lightblue"),
  addEllipses = TRUE,
  ellipse.type = "convex"
)
```

```{r}
fviz_pca_ind(
  pca,
  geom.ind = "point",
  axes = c(1,2),
  pointsize = 2,
  col.ind = estimated_sex,
  palette = c("hotpink", "lightblue"),
  addEllipses = TRUE,
  ellipse.type = "convex"
)
```

It's clear how the third component captures an information about the samples estimated gender the second and first components are not predicting.



# Identification of of differentially expressed genes (DEGs)

Differential expression analysis was performed using the DESeq2 statistical framework to identify genes showing significant expression changes between the different biological conditions.  
Since the dataset includes three groups (`control`, `glioma_CE`, and `glioma_NE`), three pairwise comparisons were carried out:

- glioma_CE vs control 
- glioma_NE vs control 
- glioma_NE vs glioma_CE  

These contrasts allow the identification of both general tumor-associated transcriptional alterations and more specific molecular differences between contrast-enhancing and non-enhancing glioma regions, providing insight into intratumoral heterogeneity.

The `results()` function was used to extract differential expression statistics from the fitted DESeq2 model for each comparison, including log2 fold changes, Wald test p-values, and adjusted p-values corrected for multiple testing.

```{r}
#resultsNames(dds)

#glioma_ce - control
res_ce = results(dds, contrast = c("class_3way", "glioma_CE", "control"))
```

```{r}
#glioma_ne - control
res_ne = results(dds, contrast = c("class_3way", "glioma_NE", "control"))
```

```{r}
res_ne_ce = results(dds, contrast = c("class_3way", "glioma_NE", "glioma_CE"))
```

Missing values were removed from the DEG result tables using `na.omit()`.  
These NA values are typically associated with genes showing extremely low expression levels or insufficient information for reliable statistical estimation. Removing them simplifies downstream analyses and visualization steps by retaining only genes with valid differential expression statistics.

```{r}
res_ne = na.omit(res_ne)
res_ce = na.omit(res_ce)
res_ne_ce = na.omit(res_ne_ce)

print(dim(res_ce))
print(dim(res_ne))
print(dim(res_ne_ce))

```

## MA plot

MA plots were generated to visualize the relationship between the mean of normalized counts and the log2 fold change across the different comparisons.  
In these plots, each point represents a gene, the x-axis corresponds to the **average normalized expression level** across samples, while the y-axis represents the **log2 fold change** between the two conditions.  
MA plots provide an overview of the global differential expression patterns, allowing the identification of genes showing substantial expression changes as well as the overall distribution of transcriptional differences between groups.

```{r}
plotMA(
  res_ce,
  ylim = c(-8, 8),
  colSig = "deeppink1",
  colNonSig = "grey58",
  main = "MA plot: glioma_CE vs control"
)
```
This first MA plot that compare glioma_CE and control samples shows a broad distribution of genes with both positive and negative log2 fold changes, indicating substantial transcriptional differences between tumor and non-tumor tissues.  

A large number of significantly differentially expressed genes are observed across a wide range of normalized expression values. Most genes remain centered around zero, reflecting genes with similar expression levels between the two conditions, while several genes display strong positive or negative fold changes, suggesting marked upregulation or downregulation in glioma_CE samples relative to controls.

The presence of genes with very large fold changes highlights profound molecular alterations associated with the tumor state.

```{r}
plotMA(
  res_ne,
  ylim = c(-8, 8),
  colSig = "deeppink1",
  colNonSig = "grey58",
  main = "MA plot: glioma_NE vs control"
)
```
Here the MA plot compare glioma_NE and control samples and also reveals widespread transcriptional differences between tumor and non-tumor tissues. A large number of genes show significant positive or negative log2 fold changes across different expression levels, indicating extensive deregulation associated with the glioma condition.  
The broad dispersion of points, especially among low- and medium-expression genes, may also reflect tissue sampling heterogeneity and possible contamination from surrounding non-tumoral brain tissue or infiltrating cell populations. Since margin glioma samples are often composed of mixed cellular environments, part of the observed variability could derive from differences in cellular composition rather than purely tumor-specific transcriptional alterations.  
Compared with the glioma_CE vs control comparison, the overall distribution appears slightly more compact, suggesting a somewhat lower degree of transcriptional divergence from control samples. Nevertheless, several genes still exhibit strong fold changes, highlighting relevant molecular alterations in the non-enhancing glioma region.

```{r}
plotMA(
  res_ne_ce,
  ylim = c(-8, 8),
  colSig = "deeppink1",
  colNonSig = "grey58",
  main = "MA plot: glioma_NE vs glioma_CE"
)
```
This last MA plot compare glioma_NE and glioma_CE samples. It shows that most genes are distributed around a log2 fold change close to zero, indicating broadly similar expression levels between the two tumor regions. Nevertheless, a subset of genes displays substantial positive or negative fold changes, suggesting the presence of region-specific transcriptional differences between contrast-enhancing and non-enhancing glioma areas.  
Part of these differences may also be influenced by tissue sampling heterogeneity and contamination from non-tumoral cell populations. In particular, non-enhancing regions are often characterized by infiltrating tumor cells intermixed with normal brain tissue, immune cells, and stromal components, which can contribute to the observed transcriptional variability.  
Compared with the tumor-versus-control comparisons, the overall distribution appears more compact, reflecting the greater biological similarity between the two glioma subtypes despite the presence of distinct molecular alterations.

## VOLCANO plot

Volcano plots were generated to simultaneously visualize the magnitude of gene expression changes and their statistical significance across the different comparisons.  
In these plots, each point represents a gene, the x-axis corresponds to the **log2 fold change** between the two conditions, while the y-axis represents the **negative logarithm of the adjusted p-value** (`-log10(padj)`).

This type of visualization allows the rapid identification of genes showing both large expression changes and high statistical significance. Genes located farther from zero on the x-axis and higher on the y-axis represent the most strongly differentially expressed genes between the compared conditions.

```{r}
#prepariamo il dataframe
res_df = as.data.frame(res_ce)

#prepariamo la colonna di significatività 
res_df$significant = ifelse(
  res_df$padj < 0.05 &
  abs(res_df$log2FoldChange) > 1,
  "Significant",
  "Not significant"
)

#volcano plot
ggplot(res_df,
       aes(x = log2FoldChange,
           y = -log10(padj),
           color = significant)) +

  geom_point(alpha = 0.6, size = 1.5) +

  scale_color_manual(values = c(
    "Not significant" = "grey58",
    "Significant" = "deeppink1"
  )) +

  theme_minimal() +

  labs(
    title = "Volcano plot glioma_CE vs control",
    x = "Log2 Fold Change",
    y = "-Log10 adjusted p-value"
  )
```
The volcano plot comparing glioma_CE and control samples highlights a large number of significantly differentially expressed genes between tumor and non-tumor tissues.  
Significant genes are distributed on both sides of the plot, indicating both upregulated and downregulated genes in glioma_CE samples relative to controls. Several genes also show very high fold changes and strong statistical significance, reflecting substantial transcriptional alterations associated with the tumor condition.  

```{r}
#prepariamo il dataframe
res_df = as.data.frame(res_ne)

#prepariamo la colonna di significatività 
res_df$significant = ifelse(
  res_df$padj < 0.05 &
  abs(res_df$log2FoldChange) > 1,
  "Significant",
  "Not significant"
)

#volcano plot
ggplot(res_df,
       aes(x = log2FoldChange,
           y = -log10(padj),
           color = significant)) +

  geom_point(alpha = 0.6, size = 1.5) +

  scale_color_manual(values = c(
    "Not significant" = "grey58",
    "Significant" = "deeppink1"
  )) +

  theme_minimal() +

  labs(
    title = "Volcano plot glioma_NE vs control",
    x = "Log2 Fold Change",
    y = "-Log10 adjusted p-value"
  )
```
This second volcano plot compare glioma_NE and control samples, it also shows extensive differential expression patterns, with numerous significantly deregulated genes across the two conditions.  
Compared with the glioma_CE comparison, the distribution appears slightly less extreme, although several genes still exhibit high fold changes and strong statistical significance, supporting the presence of marked molecular alterations in the non-enhancing glioma region.  
The less pronounced separation between conditions may partially reflect tissue sampling heterogeneity and contamination from surrounding non-tumoral brain tissue. Since non-enhancing glioma regions often contain infiltrating tumor cells intermixed with normal cellular components, part of the observed transcriptional profile could derive from differences in cellular composition.

```{r}
#prepariamo il dataframe
res_df = as.data.frame(res_ne_ce)

#prepariamo la colonna di significatività 
res_df$significant = ifelse(
  res_df$padj < 0.05 &
  abs(res_df$log2FoldChange) > 1,
  "Significant",
  "Not significant"
)

#volcano plot
ggplot(res_df,
       aes(x = log2FoldChange,
           y = -log10(padj),
           color = significant)) +

  geom_point(alpha = 0.6, size = 1.5) +

  scale_color_manual(values = c(
    "Not significant" = "grey58",
    "Significant" = "deeppink1"
  )) +

  theme_minimal() +

  labs(
    title = "Volcano plot glioma_NE vs glioma_CE",
    x = "Log2 Fold Change",
    y = "-Log10 adjusted p-value"
  )
```
The volcano plot highlights numerous significantly differentially expressed genes between glioma_NE and glioma_CE samples.  
Significant genes are distributed on both sides of the plot, indicating the presence of genes specifically upregulated in each tumor region. Most genes display moderate fold changes around zero, reflecting the overall biological similarity between the two glioma subtypes.  
Nevertheless, several genes show strong expression differences and high statistical significance, supporting the presence of region-specific transcriptional programs within the tumor microenvironment.  
Part of the observed variability may also derive from tissue sampling heterogeneity and contamination from non-tumoral cellular components. In particular, non-enhancing regions often contain infiltrating tumor cells intermixed with normal brain tissue, immune cells, and stromal populations, which can influence the transcriptional profile and contribute to the moderate separation observed between the two regions.

## Top genes

Differentially expressed genes (DEGs) were filtered by selecting genes with an adjusted p-value (`padj`) lower than **0.05**.  
The number of significant DEGs was then evaluated for each comparison in order to quantify the extent of transcriptional differences between the analyzed conditions.

For each comparison, significant genes were ranked according to adjusted p-value, and the top 20 most significant DEGs were selected for further inspection.  
Since gene identifiers were initially reported as Ensembl IDs, an additional annotation step was performed to convert them into gene symbols, allowing a clearer biological interpretation of the results.

```{r}
#filtering DEGs 
degs_ce= res_ce[res_ce$padj < 0.05, ]
cat("In glioma_ce against control there are",nrow(degs_ce),"DEGs \n")

degs_ne= res_ne[res_ne$padj < 0.05, ]
cat("In glioma_ne against control there are",nrow(degs_ne),"DEGs\n")

degs_ne_ce= res_ne_ce[res_ne_ce$padj < 0.05, ]
cat("In glioma_ne against glioma_ce there are",nrow(degs_ne_ce),"DEGs")
```

```{r}
# ordering the most significative

for (v in c("_ce", "_ne", "_ne_ce")) {  
  nome_degs = paste0("degs",v)
  nome_ordered = paste0("res_ordered",v)
  nome_top = paste0("top_deg",v)
  
  assign(nome_ordered, get(nome_degs)[order(get(nome_degs)$padj), ])
  assign(nome_top, head(get(nome_ordered),20))
  
  # ID conversion 
  temp_df <- get(nome_top)
  temp_df$Gene_clean = gsub("\\..*", "", rownames(temp_df))
  
  
  #riusiamo mart x conversione
  gene_symbols <- getBM(
    attributes = c(
      "ensembl_gene_id",
      "external_gene_name"
    ),
    filters = "ensembl_gene_id",
    values = temp_df$Gene_clean,
    mart = mart
    )
    
  temp_df <- data.frame(Ensembl_ID = rownames(temp_df),temp_df)
    
    temp_df <- temp_df |>
    left_join(
      gene_symbols,
      by = c("Gene_clean" = "ensembl_gene_id")
    )
    
    temp_df$external_gene_name[
  is.na(temp_df$external_gene_name) |
  temp_df$external_gene_name == ""
] <- "nameless_gene"
    
  #clean teable
    
  temp_df = temp_df[, c(
      "external_gene_name",
      "log2FoldChange",
      "Ensembl_ID"
    )]
    
  colnames(temp_df) = c(
      "Gene",
      "Log2FoldChange"
    )
    
  assign(nome_top, temp_df)
  
  
  }
 
knitr::kable(caption = "TOP 20 DEGs between glioma_ce vs control",top_deg_ce[,1:2])
```
The top differentially expressed genes identified in the glioma_CE versus control comparison reveal strong transcriptional alterations associated with the tumor condition.  
Several highly upregulated genes belong to the *HOXC* family, including *HOXC9*, *HOXC11*, and *HOXC13*, which are known developmental transcription factors frequently associated with oncogenic processes, cellular proliferation, and tumor progression.

In addition, genes such as *PITX2* and multiple long non-coding RNAs (HOXC-AS2, IL21-AS1, and LINC family members) suggest the activation of regulatory pathways involved in epigenetic modulation and transcriptional reprogramming. The very high log2 fold change values observed for these genes indicate substantial molecular divergence between contrast-enhancing glioma tissue and normal control samples, supporting the presence of a strongly altered tumor-associated transcriptional profile.  

```{r}
knitr::kable(caption = "TOP 20 DEGs between glioma_ne vs control",top_deg_ne[,1:2])
```
The top differentially expressed genes identified in the glioma_NE versus control comparison also reveal marked transcriptional alterations associated with the tumor condition.  
Several strongly upregulated genes belong to the *HOXC* family, including *HOXC9*, *HOXC11*, and *HOXC13*, together with regulatory non-coding transcripts such as *HOXC-AS2* and *IL21-AS1*. These genes are commonly associated with developmental programs, transcriptional regulation, and oncogenic processes.  
The presence of genes such as *PITX2* and multiple long non-coding RNAs further supports the activation of tumor-related regulatory pathways in the non-enhancing glioma region.  
Although the overall expression profile appears broadly similar to that observed in the contrast-enhancing region, the slightly lower fold changes may reflect the greater cellular heterogeneity of glioma_NE samples, which often contain infiltrating tumor cells intermixed with non-tumoral brain tissue.

Overall, these findings indicate that the non-enhancing region still retains a strongly altered molecular profile compared with normal control tissue, despite the presence of potential tissue composition differences.

```{r}
knitr::kable(caption = "TOP 20 DEGs between glioma_ne vs glioma_ce",top_deg_ne_ce[,1:2])
```
The top differentially expressed genes between glioma_NE and glioma_CE reveal important regional transcriptional differences within the tumor microenvironment. Several highly expressed genes, including *SYP*, *CAMK2A*, *NRGN*, *PRKCG*, *SLC6A7*, and *TBR1*, are strongly associated with neuronal signaling, synaptic activity, and neurodevelopmental functions.  
The enrichment of neuronal-related genes in glioma_NE samples may reflect the infiltrative nature of non-enhancing glioma regions, which are often characterized by tumor cells intermixed with surrounding normal brain tissue. This finding suggests that part of the observed transcriptional profile could derive from tissue composition heterogeneity and contamination from non-tumoral neural cell populations.  
In addition, genes such as *ARHGAP44* and *RAPGEF4*, involved in intracellular signaling and cytoskeletal regulation, may indicate molecular mechanisms associated with tumor cell plasticity and regional adaptation.  
Overall, these results support the presence of biologically distinct transcriptional programs between contrast-enhancing and non-enhancing glioma regions.

## HEATMAP

A heatmap was generated using the top differentially expressed genes identified across all pairwise comparisons in order to visualize global expression patterns and sample clustering among the analyzed conditions.  
Expression values were extracted from the variance stabilized transformed (VST) dataset and scaled by row to facilitate the comparison of relative gene expression across samples.

```{r}
all_top <- bind_rows(
  top_deg_ce,
  top_deg_ne,
  top_deg_ne_ce
)

top_genes = distinct(all_top, ...3, .keep_all = TRUE)

#estraiamo i dati normalizzati
mat = assay(vsd)[top_genes$...3, ]

#annotiamo i campioni
annotation_col = data.frame(
  Condition = colData(SRP)$class_3way
)

rownames(annotation_col) = colnames(mat)
rownames(mat) <- top_genes$Gene

my_colors = list(Condition = Class_pal3way)

#heatmap
pheatmap(
  mat,
  scale = "row",
  annotation_col = annotation_col,
  show_rownames = TRUE,
  clustering_method = "complete",
  color = continuous_pal3,
  main = "Top differentially expressed genes",
  annotation_colors = my_colors,
  show_colnames = FALSE
)
```
The heatmap highlights distinct expression patterns across the three analyzed conditions, with samples showing a tendency to cluster according to their biological group.  
In particular, control samples appear clearly separated from glioma samples, reflecting the strong transcriptional differences between tumor and non-tumor tissues.  
Glioma_CE and glioma_NE samples show partially overlapping but still distinguishable expression profiles, supporting the presence of both shared tumor-related signatures and region-specific molecular alterations. The partial overlap between the two glioma regions may also reflect tissue heterogeneity and contamination from non-tumoral cellular components, especially in non-enhancing samples where infiltrating tumor cells are often intermixed with surrounding brain tissue.  
Overall, the observed clustering patterns confirm that the selected top DEGs capture biologically relevant transcriptional variability across the dataset.

# Gene set analysis



```{r}
#dim(degs_ce)
#dim(degs_ne)
#dim(degs_ne_ce)
```
```{r}
for (v in c("_ce", "_ne", "_ne_ce")) {
  nome_res = paste0("res",v)
  nome_degs = paste0("degs",v)
  
  nome_kk =paste0("kk",v)
  
  universe_genes <- row.names(get(nome_res))
  deg_genes <- row.names(get(nome_degs))
  
  universe_genes <- gsub("\\..*", "", universe_genes)
  deg_genes <- gsub("\\..*", "", deg_genes)
  
  
  universo.entrez <- mapIds(org.Hs.eg.db, # genome wide annotation for Human that contains Entrez Gene identifiers
                    keys = universe_genes, #Column containing Ensembl gene ids
                    column = "ENTREZID",
                    keytype = "ENSEMBL")  

  sign.entrez <- mapIds(org.Hs.eg.db,
                    keys = deg_genes, #Column containing Ensembl gene ids
                    column = "ENTREZID",
                    keytype = "ENSEMBL")
  
  universo.entrez <- as.character(universo.entrez)
  sign.entrez <- as.character(sign.entrez)
  
  
  assign(nome_kk, enrichKEGG(gene = sign.entrez,
                 universe = universo.entrez,
                 organism = 'hsa',
                 keyType = 'ncbi-geneid',
                 pvalueCutoff = 0.05,
                 pAdjustMethod = "BH"))
  
  print(dim(get(nome_degs)))
  print(dim(get(nome_res)))
  
}


```

```{r}
kk_ce@result[1:10,]
kk_ne@result[1:10,]
kk_ne_ce@result[1:10,]
```
**QUESTE TABELLE SONO ORDINATE X GENI + IMPORTANTI O NO?**
```{r}
barplot(kk_ce)
dotplot(kk_ne)

nrow(as.data.frame(kk_ce))
```

**CI SONO TROPPI CAZZO DI DEGS, NON Puo ESSERE DEG IL 60% DEL TRASCRITTOMA UMANO....**


# ALTRO è PUNTI BONUSSSS

# Conclusions






































