---
title: "Molecular profiling of glioblastoma: spatial comparison against control brain tissue"
author: "Chiara Falleroni & Luca Marzilli"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: False
    number_sections: true
    fig_width: 8
    fig_height: 6
    highlight: tango
    df_print: kable
  html_document: 
    toc: true
    toc_float: true
    theme: cosmo
    number_sections: true
    fig_width: 8
    fig_height: 6
editor_options: 
  markdown: 
    # chiedi a chiara se le piaciono gli output inline.
    wrap: 80
    # chunk_output_type: console 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
```


\newpage
\tableofcontents
\newpage

```{r}
library(omicsdata)
library(SummarizedExperiment)
library(tidyverse)
```

# Biological context and study objectives 

# Experimental design 

# Data and preprocessing 

## Data loading and a brief look at it

```{r}
if (file.exists("SRP044668.rds")) {
  SRP = readRDS("SRP044668.rds")
} else {
  stop("dear user: the necessary data is not present in the current working directory or 
        does not have the right name, please check.")
}
```

Let's give a look at the data's structure. 

```{r}
#str(SRP)
dim(SRP)
assays(SRP)
```

## Data subsetting

We are now going to subset the data coherently with our goals.

```{r}
sample_info = colData(SRP)$sra.sample_attributes

# Using new columns in the ColData I classify the columns (samples) of the 
# dataset based on the type.

# 3-way classification

colData(SRP)$class_3way =
  case_when(grepl("brain tissue", sample_info)~"control",
            grepl("contrast-enhancing", sample_info)~"glioma_CE",
            grepl("non-enhancing", sample_info)~"glioma_NE")

# 2-way classification

colData(SRP)$class_2way =
  case_when(grepl("brain tissue", sample_info)~"control",
            grepl("glioma", sample_info)~"glioma")

# FACTORIALIZATION (really helpful in graphs and other procedures)

colData(SRP)$class_3way = factor(colData(SRP)$class_3way, 
                                  levels = c("control", "glioma_CE", "glioma_NE"))

colData(SRP)$class_2way = factor(colData(SRP)$class_2way, 
                                  levels = c("control", "glioma"))
```

## Unexpressed genes removal

As a team, we decide to exclude genes with expression levels at 0 or really low levels in all samples here, before the EDA, and not in the quality control paragraph. We do this because we believe this row filtering is not a quality control but a basic definition of the rows we consider "data".

```{r}
table(rowSums(assay(SRP, "raw_counts")) == 0)
table(rowSums(assay(SRP, "raw_counts")) < 15)

table(is.na(assay(SRP, "raw_counts"))) #good!
table(duplicated(rownames(SRP))) #good!

SRP = SRP[rowSums(assay(SRP, "raw_counts")) > 15, ]

```


# Pre-normalization EDA 

# Data normalization and quality control

# Post-normalization EDA

# Identification of DEGs

# Gene set analysis

# ALTRO è PUNTI BONUSSSS

# Conclusions







