---
title: "Molecular profiling of glioblastoma: spatial comparison against control brain tissue"
author: "Chiara Falleroni & Luca Marzilli"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: False
    number_sections: true
    fig_width: 8
    fig_height: 6
    highlight: tango
    df_print: kable
  html_document: 
    toc: true
    toc_float: true
    theme: cosmo
    number_sections: true
    fig_width: 8
    fig_height: 6
editor_options: 
  markdown: 
    # chiedi a chiara se le piaciono gli output inline.
    wrap: 80
    # chunk_output_type: console 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
```


\newpage
\tableofcontents
\newpage

```{r}
library(RColorBrewer)
library(omicsdata)
library(SummarizedExperiment)
library(tidyverse)
library(tibble)
library(edgeR)
library(limma)
library(DESeq2)
library(EDASeq)
#install.packages("factoextra")
library(factoextra)
```

```{r}
Class_pal2way = c("control"="turquoise4", "glioma"="firebrick2")
Class_pal3way = c("control"="turquoise4", "glioma_CE"="darkred", "glioma_NE"="salmon2")
continuous_pal2 = colorRampPalette(c("plum1","purple4"))(100)
continuous_pal3 = colorRampPalette(c("springgreen4", "antiquewhite1", "orangered2"))(100)
```

# Biological context and study objectives 

This work is based on the study by Gill et al., 2014, which investigates the molecular and transcriptional landscape of glioblastoma using RNA sequencing approaches. The article provides a comprehensive framework for understanding the complexity of this tumor by integrating high-throughput sequencing data with spatial and biological context, offering valuable insights into the heterogeneity and cellular composition of glioblastoma.  

In this article, ***glioblastoma (GBM)*** is described as the most aggressive primary brain tumor and is characterized by extensive intratumoral heterogeneity and diffuse infiltration into surrounding brain tissue.  

Due to this infiltrative behavior, complete surgical resection is not possible, and residual tumor cells located in the margins of the tumor inevitably contribute to disease recurrence. 
Recent transcriptomic studies have shown that glioblastoma can be classified into distinct molecular subtypes, including proneural, classical, and mesenchymal.  

However, most of these classifications are based on samples derived from the **contrast-enhancing (CE)** tumor core, which corresponds to regions that appear enhanced in MRI scans due to the presence of a disrupted blood-brain barrier and are typically enriched in highly proliferative tumor cells. This focus on the tumor core may overlook critical aspects of tumor biology, particularly those related to invasion and interaction with the surrounding microenvironment.  

In contrast, **non-enhancing (NE)** tumor regions refer to the infiltrative margins of the tumor that do not show contrast enhancement in imaging and contain a heterogeneous mixture of neoplastic cells and normal brain cells, such as neurons and glial cells. These regions are thought to play a crucial role in tumor progression and recurrence, as they harbor residual tumor cells that evade surgical resection and may exhibit distinct transcriptional programs compared to the tumor core.  

In this study, we analyze RNA-seq data from glioma and non-neoplastic brain tissues to investigate differences in gene expression profiles. Specifically, we aim to compare tumor samples with normal brain tissue and to further explore differences between CE and NE tumor regions. This allows us to assess how cellular composition and tumor heterogeneity influence gene expression patterns.

\vspace{2cm}

The main objectives of this analysis are:  

- to identify genes differentially expressed between glioma and non-neoplastic tissues;  
- to evaluate the extent to which tumor samples differ from normal brain tissue at the transcriptomic level;  
- to investigate how different tumor regions (CE vs NE) contribute to variability in gene expression;  
- to interpret the results in light of cellular heterogeneity and tumor microenvironment.



# Experimental design 

## Data provenance: original article's methods

### Physical collection of samples and replicates

The researchers analyzed 69 adult patients who were presenting for open surgical resection of a glioma. They also collected 17 samples of healthy (non neoplastic) brain tissue from 11 patients undergoing surgery for other reasons (such as hydrocephalus or epilepsy) to use as a baseline for comparison. 

Before beginning to remove the main bulk of the tumor (surgical debulking), the neurosurgeons took small fragments of tissue (measuring approximately 1.0 x 0.5 x 0.5 cm). To know exactly where they were taking the samples from, they used an intraoperative neuro-navigation system (Brainlab) based on MRI scans. They specifically targeted two areas: the "core" of the tumor (the contrast-enhancing area, called "CE") and the infiltrative margins of the tumor (the non-enhancing area, visible on FLAIR sequences, called "NE").

Once taken, directly in the operating room, each piece of tissue was divided in half: one piece was immediately flash-frozen in liquid nitrogen (to preserve the RNA unaltered), while the other piece was immersed in 10% formalin to "fix" and preserve the tissue structure for microscopic analysis.

### Histological analysis 

The formalin-fixed samples were cut into very thin slices, stained, and analyzed under a microscope to visually evaluate cellular density, the presence of abnormalities (like necrosis or blood vessel proliferation), and to physically mark and count the remaining entrapped healthy neurons (using a marker called NeuN).

### RNA-sequencing

Genome-wide expression profiles were generated by performing RNA-sequencing on a total of 92 flash-frozen samples. These included 39 from the tumor core (CE), 36 from the infiltrative margins (NE), and 17 from healthy brain tissue (NB).

For each sample, sequencing "libraries" were prepared using the Illumina TruSeq v2 RNA-Seq kit. The actual sequencing was performed on an Illumina HiSeq 2000 machine, which produced between 15 and 30 million "single-end, 100-base reads" per sample.

Alignment was performed via the software Tophat 2, and expression levels were calculated with Cufflinks 2, using the FPKM metric (Fragments Per Kilobase per Million mapped reads).

### Original author's bioinformatic pipeline

Using the RNA-seq data, they calculated a mathematical correlation (Spearman) between the genetic profiles of their samples and those of a massive public reference database (TCGA). This was used to classify each region of the tumor into one of the known molecular subtypes (Proneural, Classical, Mesenchymal, or Neural).

Subsequently, the authors conducted a **Computational deconvolution**: a mathematical practice performed in order to "untangle" the signal from tumor margins; in fact this samples where often a complex mixture (see the previous paragraph) of tumor and non-tumor brain cells.  
Before the deconvolution the most of the NE tumor samples were assigned to the "neural" TCGA expression profile; but this stopped being the case after mathematical unmixture.   
The authors therefore implicitly demostrated how the neural subtype was not an actual expression profile for the formerly analyzed gliomas, but an artifact caused by cellular contamination.  

## Dataset structure

The gene expression data related to the study were imported into the analysis environment as a SummarizedExperiment object (originating from the SRP044668 project), a standard data structure in Bioconductor designed to securely coordinate expression matrices with their respective clinical and technical metadata. 

The dataset consists of a single expression matrix, named raw_counts, which contains the unnormalized raw read counts. This matrix is substantial in size: it contains expression values for 63,856 genes (annotated for Homo sapiens) measured across a set of 94 biological samples.  
Alongside the data matrix, the object integrates an extensive annotation and metadata table (colData) consisting of 175 variables for each sample.

# Data and preprocessing 

## Data loading and a brief look at it

```{r}
if (file.exists("SRP044668.rds")) {
  SRP = readRDS("SRP044668.rds")
} else {
  stop("dear user: the necessary data is not present in the current working directory or 
        does not have the right name, please check.")
}
```

Let's give a look at the data's structure. 

```{r}
#str(SRP)
dim(SRP)
assays(SRP)
```

## Data subsetting

We are now going to subset the data coherently with our goals.

```{r}
#sample_info = colData(SRP)$sra.sample_attributes 

# re-ordering that annoying last observation with the other glioma_NE
nuovo_ordine <- c(94, 1:93)
SRP = SRP[, nuovo_ordine]

sample_info = colData(SRP)$sra.sample_attributes

# Using new columns in the ColData I classify the columns (samples) of the 
# dataset based on the type.

# 3-way classification

colData(SRP)$class_3way =
  case_when(grepl("brain tissue", sample_info)~"control",
            grepl("contrast-enhancing", sample_info)~"glioma_CE",
            grepl("non-enhancing", sample_info)~"glioma_NE")

# 2-way classification

colData(SRP)$class_2way =
  case_when(grepl("brain tissue", sample_info)~"control",
            grepl("glioma", sample_info)~"glioma")

# FACTORIALIZATION (necessary for graphs and other procedures)

colData(SRP)$class_3way = factor(colData(SRP)$class_3way, 
                                  levels = c("control", "glioma_CE", "glioma_NE"))

colData(SRP)$class_2way = factor(colData(SRP)$class_2way, 
                                  levels = c("control", "glioma"))
```

## Unexpressed genes removal

As a team, we decide to exclude genes with expression levels at 0 or extremely low levels in all samples here, before the EDA, and not in the quality control paragraph. We do this because we believe this row filtering is not a quality control but a basic definition of the rows we consider "data". We also decide to use the FilterByExpr (from edgeR) function, a much more statistically reliable tool.

```{r}
table(rowSums(assay(SRP, "raw_counts")) == 0)
table(rowSums(assay(SRP, "raw_counts")) < 15)

table(is.na(assay(SRP, "raw_counts"))) #good!
table(duplicated(rownames(SRP))) #good!

dim(SRP)
keep = filterByExpr(assay(SRP, "raw_counts"), group = SRP$class_3way)
SRP = SRP[keep, ]
dim(SRP)

```


# Pre-normalization EDA 

Now, before proceeding, we will use some graphical tools to investigate the general nature of our counts. This analysis is going to be helpful in guiding us through the right choices in the next section: data normalization and quality control.

## Analysis of the genetic expression distribution in the samples

```{r}

meta_df = data.frame(Sample=colnames(SRP), Class2way=SRP$class_2way, Class3way=SRP$class_3way)

dfCounts_long = as.data.frame(assay(SRP, "raw_counts")) |>
  rownames_to_column(var="Gene") |>
  pivot_longer(cols=-Gene, names_to = "Sample", values_to = "Expression")

dfCounts_long = dfCounts_long |> left_join(meta_df, by = "Sample")

somechosensamples = c("SRR1521363", "SRR1521352", "SRR1521353", "SRR1521354", 
                      "SRR1521355", "SRR1521356", "SRR1521357", "SRR1521358",
                      "SRR1521392", "SRR1521393", "SRR1521394", "SRR1521395", 
                      "SRR1521396", "SRR1521397", "SRR1521398", "SRR1521399",
                      "SRR1521440", "SRR1521441", "SRR1521442", "SRR1521443", 
                      "SRR1521444", "SRR1521445")

dfCounts_long |>
  filter(Sample %in% somechosensamples) |>
  ggplot( aes(x=Sample, y=Expression, fill = Class2way)) +
  geom_boxplot() +
  scale_fill_manual(values=Class_pal2way) +
  theme(
    axis.text.x = element_blank(),  
    axis.ticks.x = element_blank(), 
    axis.title.x = element_blank()  
  )

  
```
This are boxplots of the expression levels of some randomly chosen samples. As seen in the laboratories, we clearly need a logarithmic transformation.

```{r}
dfCounts_long$Expression = log1p(dfCounts_long$Expression)

dfCounts_long |>
  filter(Sample %in% somechosensamples) |>
  ggplot( aes(x=Sample, y=Expression, fill = Class2way)) +
  geom_boxplot() +
  scale_fill_manual(values=Class_pal2way) +
  theme(
    axis.text.x = element_blank(),  
    axis.ticks.x = element_blank(), 
    axis.title.x = element_blank())

# creation of a new log1p assay in the SRP summarized experiment object
assays(SRP)$logcounts <- log1p(assay(SRP, "raw_counts"))

```
Now the distributions are much better.

Here we show the distribution densities conditioned on the sample class of the expressions.

```{r}
dfCounts_long |>
  ggplot(aes(x=Expression, fill=Class2way, colour = Class2way)) +
        geom_density(alpha=0.4) +
        scale_fill_manual(values = Class_pal2way) +
        scale_color_manual(values = Class_pal2way)

dfCounts_long |>
  ggplot(aes(x=Expression, fill=Class2way, colour = Class3way)) +
        geom_density(alpha=0.4) +
        scale_fill_manual(values = Class_pal3way) +
        scale_color_manual(values = Class_pal3way)
```
## MD, MA and RLE plots

### MA

```{r}
limma::plotMA(assay(SRP, "logcounts"))
```

Here the limma's plotMA() function is working as a quality control tool. It's confronting the first sample with the mean of the others, and the asymmetries clearly comunicate us the need for normalization.

### MD

```{r}
EDASeq::MDPlot(assay(SRP, "raw_counts"), c(6, 7))
```

### RLE

```{r}
plotRLE(assay(SRP,"raw_counts"), outline = FALSE, las = 2,
        ylab = "RLE", main = "RLE of raw counts", col = Class_pal3way)
```

Both the MD and RLE plots strengthen what was already visible in the MA: the data need normalization.

## PCA

```{r}
pca = prcomp(t(assay(SRP, "logcounts")))
screeplot(pca)

fviz_pca_ind(pca,
             geom.ind = "point",       
             pointsize = 1,            
             col.ind = SRP$class_3way,      
             palette = Class_pal3way,       
             addEllipses = TRUE,       
             ellipse.type = "convex")
```

Finally, we here performed and plotted a pca on the un-normalized data. 

## GC content bias evaluation  
GC content can introduce biases in library preparation and sequencing, potentially affecting the quantification of gene expression.
It's important to check GC content bias for eventually filtering out sequences.  
First, we need to retrieve the amount of *GC* for each gene. To do this we use the **rowRanges** row variables and the **getGeneLengthAndGCContent** function of the *EDASeq* package.

```{r}
#Gene ID extraction and annotation retrieval
gene_ids = as.character(rowRanges(SRP)$gene_id)
gene_ids = sub("\\..*$", "", gene_ids)

gc_info = getGeneLengthAndGCContent(
  id = gene_ids,
  org = "hg38",
  mode = "org.db"
)
```

Gene identifiers were extracted from the dataset and processed to remove version numbers. GC content and gene length information were then retrieved using EDASeq annotation functions based on the hg38 reference genome.

```{r}
#Data cleaning and alignment
gc_info = gc_info[!is.na(gc_info[, 2]), ]

valid_genes = gene_ids %in% rownames(gc_info)

SRP_filtered = SRP[valid_genes, ]
gene_ids = gene_ids[valid_genes]
```

Genes lacking GC content information were removed.  
The dataset was then aligned with the annotation data to retain only genes with available GC information, ensuring consistency between expression data and gene features.

```{r}
#Count matrix preparation
counts_matrix = assay(SRP_filtered, "raw_counts")

rownames(counts_matrix) = gene_ids
```

The raw count matrix was extracted and gene identifiers were standardized to ensure consistency with the annotation dataset.

```{r}
#Removal of duplicated genes
dup_genes = duplicated(rownames(counts_matrix))

counts_matrix = counts_matrix[!dup_genes, ]
gc_info = gc_info[!dup_genes, ]
SRP_filtered = SRP_filtered[!dup_genes, ]
```

Duplicated gene identifiers, arising from multiple transcript mappings, were removed to ensure a one-to-one correspondence between genes and expression values.

```{r}
#Construction of SeqExpressionSet
seq_set = newSeqExpressionSet(
  counts = counts_matrix,
  phenoData = AnnotatedDataFrame(
    data.frame(
      condition = SRP_filtered$class_3way,
      row.names = colnames(counts_matrix)
    )
  ),
  featureData = AnnotatedDataFrame(
    data.frame(
      gc_content = gc_info[, 2],
      gene_length = gc_info[, 1]
    )
  )
)
```

A SeqExpressionSet object was constructed to integrate expression data with sample metadata and gene-level features, including GC content and gene length, enabling downstream bias analysis.

```{r}
#GC bias plot
biasPlot(seq_set, "gc_content", log = TRUE)
```
The GC bias plot shows a clear non-linear relationship between gene expression and GC content across all samples. Expression levels increase at intermediate GC values and decrease at both low and high GC content, indicating the presence of GC-dependent technical bias. The consistency of this pattern across samples suggests that the bias is systematic rather than sample-specific.

GC content bias was evaluated as part of the quality assessment to investigate potential gene-specific technical effects. Although a GC-dependent trend was observed, normalization was subsequently performed using DESeq2, which accounts for major sources of between-sample variability. Therefore, GC bias was not explicitly corrected but considered in the interpretation of the results.


# Data normalization and quality control

Although GC-dependent bias was observed during exploratory analysis, normalization was performed using DESeq2, which accounts for major sources of between-sample variability. Additional within-sample normalization was not applied to maintain consistency with the statistical modeling framework.


## Data extraction and DESeq2 dataset construction

```{r}
counts_matrix = assay(SRP, "raw_counts")

colData(SRP)$class_2way = factor(colData(SRP)$class_2way)

metadata = as.data.frame(colData(SRP))


dds = DESeqDataSetFromMatrix(
  countData = counts_matrix,
  colData = metadata,
  design = ~ class_2way
)

dds = dds[rowSums(counts(dds)) > 10, ]
```

A DESeq2 dataset object was created by combining the raw count matrix with sample metadata and specifying the experimental design based on the class_2way variable.  
A light additional filtering step was applied to remove residual low-count genes and ensure robustness in downstream analysis.

## DESeq2 normalization

```{r}
dds = DESeq(dds)
```

Normalization was carried out within the DESeq2 framework through the estimation of size factors, correcting for differences in sequencing depth and compositional biases across samples.

## Variance stabilizing transformation (VST)
```{r}
vsd = vst(dds, blind = FALSE)
```

A variance stabilizing transformation (VST) was then applied to reduce the dependence of variance on mean expression levels, making the data more suitable for exploratory analyses such as PCA and clustering.


# Post-normalization EDA

Post-normalization quality control is perform to assess the effectiveness of the normalization procedure. In particular, boxplots and Relative Log Expression (RLE) plots are used to evaluate the distribution of expression values across samples and to verify the reduction of technical variability. 
These visualizations allow us to determine whether samples are comparable and suitable for downstream analyses.

## Boxplot

```{r}
group_colors = ifelse(colData(SRP)$class_2way == "glioma", "hotpink", "lightblue")

boxplot(assay(vsd),
        outline = FALSE,
        las = 2,
        col=group_colors,
        ylab = "Expression (VST)",
        main = "Boxplot after normalization")
```
Boxplots of variance-stabilized data show highly consistent expression distributions across samples after normalization.  
The alignment of medians and the comparable spread of values indicate that technical variability has been effectively reduced, making the samples suitable for downstream comparative analysis.

## RLE plot 

```{r}
group_colours = ifelse(colData(SRP)$class_2way == "glioma", "orange", "violet")

plotRLE(assay(vsd),
        outline = FALSE,
        las = 2,
        col=group_colours,
        ylab = "RLE",
        main = "RLE after normalization")
```
The RLE plot shows values centered around zero with a consistent spread across samples, indicating that normalization effectively reduced technical variability.
A small number of samples display slight deviations from the overall trend, but these remain limited and do not suggest the presence of strong outliers or residual technical bias. Overall, the data appear well-balanced and suitable for downstream analysis.

Now we perform Principal Component Analysis (PCA) on variance-stabilized data to investigate the main sources of variation within the dataset and to evaluate the separation between biological groups. This analysis enables visualization of sample clustering patterns and helps assess whether normalization preserves biologically meaningful differences.

## PCA

```{r}
pca = prcomp(t(assay(vsd)))

plot(pca$x[,1], pca$x[,2],
     col = ifelse(colData(SRP)$class_2way == "glioma", "hotpink", "lightblue"),
     pch = 19,
     xlab = "PC1", ylab = "PC2")
```
The PCA plot shows a partial separation between glioma and control samples along the first principal component, indicating that biological differences contribute significantly to the overall variability.  
Control samples tend to cluster more tightly, while glioma samples display greater dispersion, reflecting higher heterogeneity within the tumor group. A certain degree of overlap is observed, likely due to the presence of non-neoplastic components within glioma samples or infiltrative tumor regions.  Overall, the distribution of samples suggests that normalization was effective and that biologically relevant variation is captured by the principal components.

# Identification of DEGs

# Gene set analysis

# ALTRO è PUNTI BONUSSSS

# Conclusions








