---
title: "Molecular profiling of glioblastoma: spatial comparison against control brain tissue"
author: "Chiara Falleroni & Luca Marzilli"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: False
    number_sections: true
    fig_width: 8
    fig_height: 6
    highlight: tango
    df_print: kable
  html_document: 
    toc: true
    toc_float: true
    theme: cosmo
    number_sections: true
    fig_width: 8
    fig_height: 6
editor_options: 
  markdown: 
    # chiedi a chiara se le piaciono gli output inline.
    wrap: 80
    # chunk_output_type: console 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
```


\newpage
\tableofcontents
\newpage

```{r}
library(RColorBrewer)
library(omicsdata)
library(SummarizedExperiment)
library(tidyverse)
library(tibble)
library(edgeR)
library(limma)
library(DESeq2)
library(EDASeq)
#install.packages("factoextra")
library(factoextra)
```

```{r}
Class_pal2way = c("control"="turquoise4", "glioma"="firebrick2")
Class_pal3way = c("control"="turquoise4", "glioma_CE"="darkred", "glioma_NE"="salmon2")
continuous_pal2 = colorRampPalette(c("plum1","purple4"))(100)
continuous_pal3 = colorRampPalette(c("springgreen4", "antiquewhite1", "orangered2"))(100)
```

# Biological context and study objectives 

This work is based on the study by Gill et al., 2014, which investigates the molecular and transcriptional landscape of glioblastoma using RNA sequencing approaches. The article provides a comprehensive framework for understanding the complexity of this tumor by integrating high-throughput sequencing data with spatial and biological context, offering valuable insights into the heterogeneity and cellular composition of glioblastoma.  
In this article, ***glioblastoma (GBM)*** is described as the most aggressive primary brain tumor and is characterized by extensive intratumoral heterogeneity and diffuse infiltration into surrounding brain tissue.  
Due to this infiltrative behavior, complete surgical resection is not possible, and residual tumor cells located in the margins of the tumor inevitably contribute to disease recurrence. 
Recent transcriptomic studies have shown that glioblastoma can be classified into distinct molecular subtypes, including proneural, classical, and mesenchymal.  
However, most of these classifications are based on samples derived from the **contrast-enhancing (CE)** tumor core, which corresponds to regions that appear enhanced in MRI scans due to the presence of a disrupted blood-brain barrier and are typically enriched in highly proliferative tumor cells. This focus on the tumor core may overlook critical aspects of tumor biology, particularly those related to invasion and interaction with the surrounding microenvironment.  
In contrast, **non-enhancing (NE)** tumor regions refer to the infiltrative margins of the tumor that do not show contrast enhancement in imaging and contain a heterogeneous mixture of neoplastic cells and normal brain cells, such as neurons and glial cells. These regions are thought to play a crucial role in tumor progression and recurrence, as they harbor residual tumor cells that evade surgical resection and may exhibit distinct transcriptional programs compared to the tumor core.  
In this study, we analyze RNA-seq data from glioma and non-neoplastic brain tissues to investigate differences in gene expression profiles. Specifically, we aim to compare tumor samples with normal brain tissue and to further explore differences between CE and NE tumor regions. This allows us to assess how cellular composition and tumor heterogeneity influence gene expression patterns.



The main objectives of this analysis are:  

- to identify genes differentially expressed between glioma and non-neoplastic tissues;  
- to evaluate the extent to which tumor samples differ from normal brain tissue at the transcriptomic level;  
- to investigate how different tumor regions (CE vs NE) contribute to variability in gene expression;  
- to interpret the results in light of cellular heterogeneity and tumor microenvironment.



# Experimental design 

# Data and preprocessing 

## Data loading and a brief look at it

```{r}
if (file.exists("SRP044668.rds")) {
  SRP = readRDS("SRP044668.rds")
} else {
  stop("dear user: the necessary data is not present in the current working directory or 
        does not have the right name, please check.")
}
```

Let's give a look at the data's structure. 

```{r}
#str(SRP)
dim(SRP)
assays(SRP)
```

## Data subsetting

We are now going to subset the data coherently with our goals.

```{r}
#sample_info = colData(SRP)$sra.sample_attributes 

# re-ordering that annoying last observation with the other glioma_NE
nuovo_ordine <- c(94, 1:93)
SRP = SRP[, nuovo_ordine]

sample_info = colData(SRP)$sra.sample_attributes

# Using new columns in the ColData I classify the columns (samples) of the 
# dataset based on the type.

# 3-way classification

colData(SRP)$class_3way =
  case_when(grepl("brain tissue", sample_info)~"control",
            grepl("contrast-enhancing", sample_info)~"glioma_CE",
            grepl("non-enhancing", sample_info)~"glioma_NE")

# 2-way classification

colData(SRP)$class_2way =
  case_when(grepl("brain tissue", sample_info)~"control",
            grepl("glioma", sample_info)~"glioma")

# FACTORIALIZATION (necessary for graphs and other procedures)

colData(SRP)$class_3way = factor(colData(SRP)$class_3way, 
                                  levels = c("control", "glioma_CE", "glioma_NE"))

colData(SRP)$class_2way = factor(colData(SRP)$class_2way, 
                                  levels = c("control", "glioma"))
```

## Unexpressed genes removal

As a team, we decide to exclude genes with expression levels at 0 or extremely low levels in all samples here, before the EDA, and not in the quality control paragraph. We do this because we believe this row filtering is not a quality control but a basic definition of the rows we consider "data". We also decide to use the FilterByExpr (from edgeR) function, a much more statistically reliable tool.

```{r}
table(rowSums(assay(SRP, "raw_counts")) == 0)
table(rowSums(assay(SRP, "raw_counts")) < 15)

table(is.na(assay(SRP, "raw_counts"))) #good!
table(duplicated(rownames(SRP))) #good!

dim(SRP)
keep = filterByExpr(assay(SRP, "raw_counts"), group = SRP$class_3way)
SRP = SRP[keep, ]
dim(SRP)

```


# Pre-normalization EDA 

Now, before proceeding, we will use some graphical tools to investigate the general nature of our counts. This analysis is going to be helpful in guiding us through the right choices in the next section: data normalization and quality control.

## Analysis of the genetic expression distribution in the samples

```{r}

meta_df = data.frame(Sample=colnames(SRP), Class2way=SRP$class_2way, Class3way=SRP$class_3way)

dfCounts_long = as.data.frame(assay(SRP, "raw_counts")) |>
  rownames_to_column(var="Gene") |>
  pivot_longer(cols=-Gene, names_to = "Sample", values_to = "Expression")

dfCounts_long = dfCounts_long |> left_join(meta_df, by = "Sample")

somechosensamples = c("SRR1521363", "SRR1521352", "SRR1521353", "SRR1521354", 
                      "SRR1521355", "SRR1521356", "SRR1521357", "SRR1521358",
                      "SRR1521392", "SRR1521393", "SRR1521394", "SRR1521395", 
                      "SRR1521396", "SRR1521397", "SRR1521398", "SRR1521399",
                      "SRR1521440", "SRR1521441", "SRR1521442", "SRR1521443", 
                      "SRR1521444", "SRR1521445")

dfCounts_long |>
  filter(Sample %in% somechosensamples) |>
  ggplot( aes(x=Sample, y=Expression, fill = Class2way)) +
  geom_boxplot() +
  scale_fill_manual(values=Class_pal2way) +
  theme(
    axis.text.x = element_blank(),  
    axis.ticks.x = element_blank(), 
    axis.title.x = element_blank()  
  )

  
```
This are boxplots of the expression levels of some randomly chosen samples. As seen in the laboratories, we clearly need a logarithmic transformation.

```{r}
dfCounts_long$Expression = log1p(dfCounts_long$Expression)

dfCounts_long |>
  filter(Sample %in% somechosensamples) |>
  ggplot( aes(x=Sample, y=Expression, fill = Class2way)) +
  geom_boxplot() +
  scale_fill_manual(values=Class_pal2way) +
  theme(
    axis.text.x = element_blank(),  
    axis.ticks.x = element_blank(), 
    axis.title.x = element_blank())

# creation of a new log1p assay in the SRP summarized experiment object
assays(SRP)$logcounts <- log1p(assay(SRP, "raw_counts"))

```
Now the distributions are much better.

Here we show the distribution densities conditioned on the sample class of the expressions.

```{r}
dfCounts_long |>
  ggplot(aes(x=Expression, fill=Class2way, colour = Class2way)) +
        geom_density(alpha=0.4) +
        scale_fill_manual(values = Class_pal2way) +
        scale_color_manual(values = Class_pal2way)

dfCounts_long |>
  ggplot(aes(x=Expression, fill=Class2way, colour = Class3way)) +
        geom_density(alpha=0.4) +
        scale_fill_manual(values = Class_pal3way) +
        scale_color_manual(values = Class_pal3way)
```
## MD, MA and RLE plots

### MA

```{r}
limma::plotMA(assay(SRP, "logcounts"))
```

Here the limma's plotMA() function is working as a quality control tool. It's confronting the first sample with the mean of the others, and the asymmetries clearly comunicate us the need for normalization.

### MD

```{r}
EDASeq::MDPlot(assay(SRP, "raw_counts"), c(6, 7))
```

### RLE

```{r}
plotRLE(assay(SRP,"raw_counts"), outline = FALSE, las = 2,
        ylab = "RLE", main = "RLE of raw counts", col = Class_pal3way)
```

Both the MD and RLE plots strengthen what was already visible in the MA: the data need normalization.

## PCA

```{r}
pca = prcomp(t(assay(SRP, "logcounts")))
screeplot(pca)

fviz_pca_ind(pca,
             geom.ind = "point",       
             pointsize = 1,            
             col.ind = SRP$class_3way,      
             palette = Class_pal3way,       
             addEllipses = TRUE,       
             ellipse.type = "convex")
```

Finally, we here performed and plotted a pca on the un-normalized data. 


# Data normalization and quality control

# Post-normalization EDA

# Identification of DEGs

# Gene set analysis

# ALTRO è PUNTI BONUSSSS

# Conclusions








